---
title: "GLM Analysis Post Hoc Tests and Visualization"
author: "Dr. Indrajit Gayari"
date: "2025-09-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE,
                      fig.align = "centre"
                      )
```

## Generalized Linear Models (GLM) in R

This R Markdown document provides a comprehensive guide to performing GLM analysis, post-hoc tests, least square means calculations, and creating publication-quality graphics.

```{r}
# Load required packages
library(tidyverse)    # Data manipulation and visualization
library(lme4)         # For mixed models (if needed)
library(car)          # For ANOVA tables
library(emmeans)      # For post-hoc tests and least square means
library(multcomp)     # For multiple comparisons
library(ggpubr)       # For publication-quality graphics
library(broom)        # For tidy model output
library(performance)  # For model diagnostics

# Set theme for publication-quality graphics
theme_set(theme_pubr())
```

## Steps for importing your own dataset for Analysis

```{r}
## Here, in this chunk ## is the comment # is the code
## First check current directory first
#getwd()
##Set working directory to a specific path
#setwd("C:/Users/YourName/Projects/MyAnalysis")

## Import data from CSV file
## Replace 'your_data.csv' with your actual file name
#data <- read.csv("your_data.csv")

##Display the first few rows of the data
#head(data) %>% kable()

##Check data structure
#str(data)

# Summary statistics
#summary(data)
```

## Example Dataset

I am create a sample dataset for demonstration purpose:

```{r}
set.seed(123)

# Create a sample dataset
n <- 120
data <- data.frame(
  treatment = rep(c("A", "B", "C", "D"), each = n/4),
  dose = rep(c("Low", "Medium", "High"), times = n/3),
  response = c(
    rpois(n/4, lambda = 10),  # Treatment A
    rpois(n/4, lambda = 15),  # Treatment B  
    rpois(n/4, lambda = 20),  # Treatment C
    rpois(n/4, lambda = 25)   # Treatment D
  ),
  covariate = rnorm(n, mean = 50, sd = 10)
)

# Convert factors
data$treatment <- factor(data$treatment)
data$dose <- factor(data$dose, levels = c("Low", "Medium", "High"))

head(data)
```

## Descriptive Statistics

```{r}
# Summary statistics
summary(data)

# Count by treatment and dose
table(data$treatment, data$dose)
```

## Fitting a GLM Model

Here I am fitting a Poisson GLM (appropriate for count data). For Gaussian data, use family = gaussian (link = "identity"). For Binomial data, use family = binomial(link = "logit")

```{r}
# Fit GLM with Poisson family
model <- glm(response ~ treatment * dose + covariate, 
             data = data, 
             family = poisson(link = "log"))

# Model summary
summary(model)

# ANOVA table
car::Anova(model, type = "III")
```

## Model Diagnostics

```{r}
# Check for overdispersion
# If residual deviance >> degrees of freedom, consider quasipoisson
dispersion_test <- model$deviance / model$df.residual
cat("Dispersion parameter:", dispersion_test, "\n")

# Residual plots
par(mfrow = c(2, 2))
plot(model)
```

# Check for influential observations

```{r}
influence_measures <- influence.measures(model)
summary(influence_measures)

# Check multicollinearity (for continuous predictors)
# vif(model) # Only for continuous predictors in this case
```

## Calculating Least-Squares Means (Estimated Marginal Means)

```{r}
# Calculate least-squares means for treatment
lsmeans_treatment <- emmeans(model, ~ treatment)
lsmeans_treatment

# For dose
lsmeans_dose <- emmeans(model, ~ dose)
lsmeans_dose

# For interaction
lsmeans_interaction <- emmeans(model, ~ treatment:dose)
lsmeans_interaction
```

## Post-hoc Tests

```{r}
# Pairwise comparisons for treatment
posthoc_treatment <- pairs(lsmeans_treatment, adjust = "tukey")
posthoc_treatment

# Pairwise comparisons for dose
posthoc_dose <- pairs(lsmeans_dose, adjust = "tukey")
posthoc_dose

# Compact letter display for treatment
cld_treatment <- cld(lsmeans_treatment, Letters = letters, adjust = "tukey")
cld_treatment

# Compact letter display for dose
cld_dose <- cld(lsmeans_dose, Letters = letters, adjust = "tukey")
cld_dose
```

## Publication-Quality Visualizations

### 1. Mean Plot with Error Bars

```{r}
# Create data frame for plotting
plot_data <- as.data.frame(lsmeans_treatment)

# Create mean plot with confidence intervals
mean_plot <- ggplot(plot_data, aes(x = treatment, y = emmean)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL), width = 0.2) +
  labs(title = "Least-Squares Means by Treatment",
       x = "Treatment", 
       y = "Estimated Marginal Mean (log scale)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

mean_plot
```

### 2. Interaction Plot

```{r}
# Get interaction means
interaction_data <- as.data.frame(lsmeans_interaction)

# Create interaction plot
interaction_plot <- ggplot(interaction_data, 
                           aes(x = dose, y = emmean, color = treatment, group = treatment)) +
  geom_line(size = 1) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL), width = 0.1) +
  labs(title = "Treatment × Dose Interaction",
       x = "Dose Level", 
       y = "Estimated Marginal Mean",
       color = "Treatment") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "top")

interaction_plot
```

### 3. Bar Plot with Significance Letters

```{r}
# Prepare data with significance letters
cld_data <- as.data.frame(cld_treatment)

# Create bar plot with significance letters
bar_plot <- ggplot(cld_data, aes(x = treatment, y = emmean, fill = treatment)) +
  geom_bar(stat = "identity", alpha = 0.7) +
  geom_errorbar(aes(ymin = emmean - SE, ymax = emmean + SE), width = 0.2) +
  geom_text(aes(label = .group, y = emmean + SE + 0.5), vjust = 0) +
  labs(title = "Treatment Means with Standard Errors",
       x = "Treatment", 
       y = "Estimated Marginal Mean",
       fill = "Treatment") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "none")

bar_plot
```

### 4. Multi-panel Publication Figure

```{r}
# Arrange multiple plots
multi_panel <- ggarrange(mean_plot, bar_plot, interaction_plot,
                         ncol = 2, nrow = 2,
                         labels = c("A", "B", "C"),
                         common.legend = FALSE)

# Add title
multi_panel <- annotate_figure(multi_panel,
                               top = text_grob("GLM Analysis Results", 
                                               color = "black", 
                                               face = "bold", 
                                               size = 14))

multi_panel
```

## Exporting Results

```{r}
##Save plots (uncomment to use)
#ggsave("mean_plot.png", mean_plot, width = 8, height = 6, dpi = 300)
#ggsave("interaction_plot.png", interaction_plot, width = 8, height = 6, dpi = 300)
#ggsave("multi_panel_plot.png", multi_panel, width = 10, height = 8, dpi = 300)

##Save results to CSV
#write.csv(as.data.frame(lsmeans_treatment), "treatment_means.csv")
#write.csv(as.data.frame(posthoc_treatment), "treatment_comparisons.csv")
```
